---
title: "Sankey Diagrams"
output: 
  cleanrmd::html_document_clean:
    theme: almond
---

```{r set global chunks, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE, message = FALSE
)
```

```{r load libraries}
library(tidyverse)
library(datasets)
library(plotly)
library(gt)
library(glue)
library(stringr)
library(DT)
library(purrr)
```

# Data Model

In this first section we discuss the structure our data frame should have to serve as the input to our Sankey/Subburst chart. The following data set shows outcome data for a collection of 100 student. The students enrolled in either `class1` or `class2`. They were assigned a grade of `A`, `B`, or `C`, and they had an outcome of `retained`, `dropped`, `stopped`, or `graduated`. 

```{r}
simple_class_df <- tibble(
  person_id = sample(1:999999, 100),
  class = sample(c('class1', 'class2'), 100, replace = TRUE),
  grade = sample(c('A', 'B', 'C'), 100, replace = TRUE),
  outcome = sample(c('retained', 'dropped', 'stopped', 'graduated'), 100, replace = TRUE)
) %>% 
  mutate(person_id = str_pad(person_id, 8, pad = '0'))

DT::datatable(simple_class_df)
```


Now compare this to a similarly structured, but more complex data set. 

```{r}
ipeds_re <- c('non_resident_alien', 'race_ethnicity_unknown', 'black_non_hispanic',
              'american_indian_or_alaskan_native', 'asian_or_pacific_islander',
              'hispanic', 'white_non_hispanic')
student_type <- c('ipeds_first_time', 'continuing', 'returning', 'transfer', 'graduate', 'ut_first_time')

complex_comparison_df <- tibble(
  person_id = sample(1:999999, 1000),
  student_type = sample(student_type, 1000, replace = TRUE),
  ipeds_race_ethn = sample(ipeds_re, 1000, replace = TRUE),
  college = sample(c('HS', 'BS', 'HU', 'SC', 'LR', 'ED'), 1000, replace = TRUE),
  gpa = sample(c('high', 'medium', 'low'), 1000, replace = TRUE),
  club = sample(c(TRUE, FALSE), 1000, replace = TRUE),
  outcome = sample(c('retained', 'dropped', 'stopped', 'graduated'), 1000, replace = TRUE)
) %>% 
  mutate(person_id = str_pad(person_id, 8, pad = '0'))

DT::datatable(complex_comparison_df)
```

If we want to produce a Sankey chart from either of these data we need to know what level each of the attribute columns should belong to. So if we are going to create a function `ut_sankey`, then we are going to need a couple input parameters. 

- Which field represents the population?
- Which level do each of the fields belong to?

For `simple_class_df` we have the following.

- population = person_id
- number of levels = 3
- level 1 = class
- level 2 = grade
- level 3 = outcome

For `complex_comparison_df` we have

- population = person_id
- number of levels = 3
- level 1 = student_type, ipeds_race_ethn
- level 2 = college, gpa
- level 3 = club
- level 4 = outcome

## The simple case

### Reducing from the complex to the simple.

For this step we want to reduce the complex case to the simple case. We do this by combining the multiple segments for our levels into a single segment. What would this look like?

```{r}
complex_reduction <- complex_comparison_df %>% 
  unite(level_1, c('student_type', 'ipeds_race_ethn'), sep = '__') %>% 
  unite(level_2, c('college', 'gpa'), sep = '__') %>% 
  unite(level_3, c('club'), sep = '__') %>% 
  unite(level_4, c('outcome'), sep = '__')
```

This is easy to do. The only complication to making this into a function is knowing how many levels there are. But we can probably get around this with `{purrr}`.

```{r}
levels_vec <- purrr::map(1:4, function(x) paste0('level', '_', x))

levels <- list(c('student_type', 'ipeds_race_ethn'),
                    c('college', 'gpa'),
                    c('club'),
                    c('outcome')
                    )

output_list <- map2(levels_vec, levels, ~ unite(complex_comparison_df, {{ .x }} , .y))
```

This is not quite what I want the purrr function to do. I want it to take just `person_id` and the level. I think we will have to write a function to do this.

```{r}
collapse_levels <- function(input_df, level_vec, levels) {
  
  output_df <- input_df %>% 
    unite({{ level_vec }}, all_of(levels), sep = '_._') %>% 
    select(person_id, {{ level_vec }})
  
  return(output_df)
}
```

```{r}
collapse_levels(complex_comparison_df, 'level_1', c('student_type', 'ipeds_race_ethn') )
```

Now lets feed this back into the our `purrr` workflow.

```{r}
output_list2 <- map2(levels_vec, levels, ~ collapse_levels(complex_comparison_df, {{ .x }} , .y))
```

So now that we have all this, what does the workflow look like?

We first define the names of the levels we are going to use and which of the fields we want to collapse. It is probably easiest if we define this as a named list.

```{r}
levels <- list(c('student_type', 'ipeds_race_ethn'),
                    c('college', 'gpa'),
                    c('club'),
                    c('outcome')
                    )

names(levels) <- purrr::map(1:4, function(x) paste0('level', '_', x))

output_list <- map2(names(levels), levels, ~ collapse_levels(complex_comparison_df, {{ .x }} , .y))

reduced_complexity_df <- output_list %>% 
  purrr::reduce(left_join, by = 'person_id')
```







